---
title: "Lab07:Machine learning and PCA"
author: "Morgan Farrell"
date: "2/8/2022"
output:
  pdf_document: default
  html_document: default
---

##Clustering with kmeans() and hclust()
We will begin by making up some data to cluster

```{r}
tmp <- c(rnorm(30, 3), rnorm(30, -3))
x<- cbind(tmp, rev(tmp))
plot(x)
```
Now we will run 'kmeans()'
```{r}
k <- kmeans(x, centers=2, nstart=20)
k
```
> What size are each cluster? 

Readout gives us 2 clusters each a size of 30, this makes sense because we told R to make us 2 clusters of 30
```{r}
k$size
```
> cluster centers?

```{r}
k$centers
```
 
>Clustering vector? 

```{r}
k$cluster
```
Clustering vector, indicates which values cluster in group 1 or group 2. The first 30 to 1 and the second 30 to 2

Plot our data with the clustering result
```{r}
plot(x, col= k$cluster) 
points(k$centers, col="blue", pch=15, cex=1.5)
```
## Hierarchical clustering 'hclust()'

```{r}
hc <- hclust(dist(x))
hc
```
Plot method for hclust()
```{r}
plot(hc)
```
The two groups show that the left has values 1-30 and the right have 31-60 again representing our two groups.


##Principal Component Analysis Data Practice
Import data of UK foods
```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

>Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```
Check your data by clicking on the x variable under data to see your data as a table or you can use 'head()' to preview the beginning
```{r}
head(x)
```

Need to fix the rownames to be the names not the numbers
```{r}
# Note how the minus indexing works
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```

Alternative row names approach to better control the table when importing
```{r}
x <- read.csv(url, row.names=1)
head(x)
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I've always used read.csv because you have the most control over your row names and column names

Next, visualize the data using a regular barplot
```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```
>Q3: Changing what optional argument in the above barplot() function results in the following plot?

First look at the help page for barplot()
Set the color to 'nrow()' giving you a color for each of your food categories
If you set beside= TRUE it will break up the row categories to put them side-by-side vs stacked
```{r}
?barplot()
mycols <- rainbow(nrow(x)) #set the rows to different colors of the rainbow store as a vector for that color
barplot(as.matrix(x), col=mycols)#use your color vector
```

>Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=mycols, pch=16)
```
The countries are written down the diagonal
The first plot is England vs Wales
Each plot are comparing two countries if they have similar values you would see a straight line indicating there is no difference or the value is equal 

>Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

N. Ireland and the other countries have a lot of differences in food consumption since they do not have graphs with straight lines. For example, the blue food group is consumed more in England, but the green food group is consumed more in Ireland. 

## PCA to Interpret Data
Next we will use PCA to interpret this data and see if it is more clear to figure out the trends

We are going to use the 'prcom()' function which expects the observations to be rows and the variables to be columns.
So we need to transpose the data.
```{r}
t(x) #t() can transpose the data frame to fit the expectations of the prcomp()
pca <- prcomp( t(x) ) 
summary(pca)
```
Notice PC1 will describe 67.44% of the variance of this data set. 
The PC2 is 29% of data. 
PC1 +PC2= 96.5% explain most of the variance of the data

```{r}
plot(pca) #Only plots % of explained variance
```

>Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
attributes(pca)
```

PCA plot or PCA score plot is PC1 vs PC2
```{r}
# Plot PC1 vs PC2
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x))
```

>Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x), col=rainbow(4))
```
```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```

```{r}
z <- summary(pca)
z$importance
```

```{r}
barplot(v, xlab="Principal Component", ylab="Percent Variation")
```
```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```
>Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,2], las=2 )
```
The PC1 vs PC2 plot show differences in several food groups mainly, Vegetables, fresh fruit, soft drinks and alcoholic beverages. These categories change the most between PC1 and PC2. Ireland eats more potatoes and drink more soft drinks. PC2 shows us that the rest of UK drink more alcohol and eat more fresh fruit. 

```{r}
## The inbuilt biplot() can be useful for small datasets 
biplot(pca)
```

## PCA of RNA-seq Data
```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```

>Q10: How many genes and samples are in this data set?

If you click on the rna.data file you can see that there are 100 genes and 5 knockouts along with 5 wild type

```{r}
pca <- prcomp(t(rna.data))
summary(pca)
```

Make a PCA score plot
```{r}
## Simple un polished plot of pc1 and pc2
plot(pca$x[,1:2], xlab="PC1", ylab="PC2")
text(pca$x[,1:2], labels=colnames(rna.data))
```

```{r}
## Variance captured per PC 
pca.var <- pca$sdev^2

## Percent variance is often more informative to look at 
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```

```{r}
barplot(pca.var.per, main="Scree Plot", 
        names.arg = paste0("PC", 1:10),
        xlab="Principal Component", ylab="Percent Variation")
```

```{r}
## A vector of colors for wt and ko samples
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))

text(pca$x[,1], pca$x[,2], labels = colnames(rna.data), pos=c(rep(4,5), rep(2,5)))
```

```{r}
library(ggplot2)

df <- as.data.frame(pca$x)

# Our first basic plot
ggplot(df) + 
  aes(PC1, PC2) + 
  geom_point()
```

```{r}
# Add a 'wt' and 'ko' "condition" column
df$samples <- colnames(rna.data) 
df$condition <- substr(colnames(rna.data),1,2)

p <- ggplot(df) + 
        aes(PC1, PC2, label=samples, col=condition) + 
        geom_label(show.legend = FALSE)
p
```
```{r}
p + labs(title="PCA of RNASeq Data",
       subtitle = "PC1 clealy seperates wild-type from knock-out samples",
       x=paste0("PC1 (", pca.var.per[1], "%)"),
       y=paste0("PC2 (", pca.var.per[2], "%)"),
       caption="BIMM143 example data") +
     theme_bw()
```
```{r}
loading_scores <- pca$rotation[,1]

## Find the top 10 measurements (genes) that contribute
## most to PC1 in either direction (+ or -)
gene_scores <- abs(loading_scores) 
gene_score_ranked <- sort(gene_scores, decreasing=TRUE)

## show the names of the top 10 genes
top_10_genes <- names(gene_score_ranked[1:10])
top_10_genes 
```

