---
title: "FindaGene_EBF1"
author: "Morgan Farrell"
date: "2/28/2022"
output: html_document
---

# Install proper packages
```{r}
library(bio3d)
#install.packages("pheatmap")
library(pheatmap)
```


# Read in the alignment that was converted in SEAVIEW from the original CLUSTALW format
```{r}
ebf1 <- read.fasta("EBF1-aln-clustalw.fst")
ebf1
```

#Identify any conserved proteins
```{r}
conv <- conserv(ebf1)
plot(conv, typ="h", xlab="Alignment Position", ylab="Conservation Score")
```


# Begin building heatmap based on conservation scores
```{r}
identity_ebf1 <- seqidentity(ebf1)
identity_ebf1
```


# Assemble heat map
```{r}
pheatmap(identity_ebf1, main="Heatmap of Predicted ROMO1", scale="none")
```

#Make consensus sequence from alignment
```{r}
cons_ebf1 <- consensus(ebf1)
cons_ebf1
```
My consensus sequence has too many gaps to use. I'll need to find the sequence with the highest identity to the others and use that

#Find highest identity sequence

```{r}
#install.packages("matrixStats")
library(matrixStats)
```

```{r}
ebf1_wo_self <- 1-identity_ebf1
ebf1_noself_matrix <- as.matrix(ebf1_wo_self)
ebf1_noself_matrix
```

```{r}
ebf1_noself_matrix$row_maximum = rowMaxs(ebf1_noself_matrix)
ebf1_noself_matrix
```

```{r}
pheatmap(1-identity_ebf1)
```

The Sea Hare had a high sequence identity let's try that as the representative sequence

#R/Bio3D to blast online, find the top 3 unique hi
```{r}
ebf1_sh_fasta <- read.fasta("SeaHare.fst")
ebf1_sh_blast <- blast.pdb(ebf1_sh_fasta)
```

```{r}
ebf1_sh_blast
```

```{r}
library(bio3d)
plot.blast(ebf1_sh_blast)
```

```{r}
#ids <- as.matrix(ebf1_sh_blast$hit.tbl[,2])
ids <- c("3MLP", "3N50_A", "3MUJ", "3MQI", "3LYR", "3MLN")
pdb.annotate(ids)
```



