---
title: "Graphs for Find a Gene"
author: "Morgan Farrell"
date: "2/22/2022"
output:
  pdf_document: default
  html_document: default
---

# Install proper packages
```{r}
library(bio3d)
#install.packages("pheatmap")
library(pheatmap)
```

# Read in the alignment that was converted in SEAVIEW from the original CLUSTALW format
```{r}
romo1 <- read.fasta("alignment-converted fasta.fst")
romo1
```
#Identify any conserved proteins
```{r}
conv <- conserv(romo1)
plot(conv, typ="h", xlab="Alignment Position", ylab="Conservation Score")
```

# Begin building heatmap based on conservation scores
```{r}
identity_romo1 <- seqidentity(romo1)
identity_romo1
```

# Assemble heat map
```{r}
pheatmap(identity_romo1, main="Heatmap of Predicted ROMO1", scale="none")
```
#Make consensus sequence from alignment
```{r}
#cons_romo1 <- consensus(romo1)
#cons_romo1
```
My consensus sequence has too many gaps to use. I'll need to find the sequence with the highest identity to the others and use that

#Find highest identity sequence

```{r}
#install.packages("matrixStats")
library(matrixStats)
```

```{r}
romo1_wo_self <- 1-identity_romo1
romo1_noself_matrix <- as.matrix(romo1_wo_self)
romo1_noself_matrix
```

```{r}
#install.packages("dplyr")
#library(dplyr)
```

```{r}
romo1_noself_matrix$row_maximum = rowMaxs(romo1_noself_matrix)
romo1_noself_matrix
```

```{r}
pheatmap(1-identity_romo1)
```
I think the brown sea anemone has the highest sequence identity

#R/Bio3D to blast online, find the top 3 unique hi
```{r}
romo1_bsa_fasta <- read.fasta("brown sea anemone.fst")
romo1_bsa_blast <- blast.pdb(romo1_bsa_fasta)
```

```{r}
romo1_bsa_blast
```


#Try again with different sequence to see if I can get higher e values

```{r}
romo1_cc_fasta <- read.fasta("cauliflower coral.fst")
romo1_cc_blast <- blast.pdb(romo1_cc_fasta)
romo1_cc_blast
```

```{r}
romo1_m_fasta <- read.fasta("mouse.fst")
romo1_m_blast <- blast.pdb(romo1_m_fasta)
romo1_m_blast
```

