---
title: "Class14_Mini Project Vaccination"
author: "Morgan Farrell"
date: "3/4/2022"
output:
  pdf_document:
    latex_engine: xelatex
---


# Import dataset for San Diego County Vaccination Status
```{r}
# Import vaccination data
vax <- read.csv("covid19vaccinesbyzipcode.csv")
head(vax)
```

# Q1. What column details the total number of people fully vaccinated? 
3 persons_fully_vaccinated

# Q2. What column details the Zip code tabulation area? 
12 zip_code_tabulation_area

Find the dates 
```{r}
#first date
vax$as_of_date[1]
#latest date
vax$as_of_date[nrow(vax)]
#install.packages("lubridate")
library(lubridate)
```

# Q3. What is the earliest date in this dataset? 
2021-01-05

# Q4. What is the latest date in this dataset?
2022-03-01

Using the skimr package to see the dataset characteristics
```{r}
#install.packages("skimr")
skimr::skim(vax)
```

# Q5. How many numeric columns are in this dataset? 
10, but Zip code is counted as numeric so it should really be 9

# Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column? 
18338

# Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)? 
```{r}
(18338/sum(vax$persons_fully_vaccinated, na.rm=TRUE))*100
```

# Q8. [Optional]: Why might this data be missing?
Because they put kids vaccination in different columns? 

# Using the package 'lubridate()'
```{r}
library(lubridate)
time_length(today()-ymd("1994-03-06"), "years") #I'm getting old
```

Store the as_of_date column as a variable so we can do math with it
```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```

# Q9. How many days have passed since the last update of the dataset? 
```{r}
today() - vax$as_of_date[1]
```

Days between the first and last vaccination
```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

# Q10. How many unique dates are in the dataset (i.e. how many different dates are detailed)? 
```{r}
length(unique(vax$as_of_date))
```


```{r}
#install.packages("zipcodeR")
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

```{r}
zip_distance('92037','92109')
```

```{r}
reverse_zipcode(c('92037', "92109") )
```

Figuring out SD counties using base R 
```{r}
sd <- vax[vax$county == "San Diego", ]
```

Using 'dplyr' package to filter the County data instead
```{r}
library(tidyverse) #this has a bunch of packages ggplot, dplyr, etc. 
```

Using the filter function
```{r}
sd <- filter(vax, county== "San Diego")
```

Using the pipe funtion to pass the variable onto the arguments
```{r}
#vax %>% filter(county == "San Diego") 
```

```{r}
head(sd, 3)
```
# Q11. How many distinct zip codes are listed for San Diego County?
```{r}
length(unique(sd$zip_code_tabulation_area))
```

```{r}
length(table(sd$zip_code_tabulation_area))
```
Using dplyr and the pipe function
```{r}
vax %>% 
  filter(county == "San Diego") %>%
  select(zip_code_tabulation_area) %>%
  unique() %>%
  nrow()
```

# Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?
92154

```{r}
#vax %>%
#  filter(county == "San Diego") %>%
#  select(age12_plus_population) %>%
#  order(decreasing = TRUE)
  
```

Base R way to answer the question
```{r}
inds <- order(sd$age12_plus_population, decreasing = TRUE)
sd[inds[1],]
```

Using dplyr and the 'arrange()' function
```{r}
head(arrange(sd, -age12_plus_population)) # the minus means sort opposite, which the default is lowest to highest population
```

# Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-02-22”?
70.53%

```{r}
#Using the pipe method, a bit complicated
sd %>%
filter(as_of_date == "2022-03-01") %>%
select(percent_of_population_fully_vaccinated) %>%
colMeans(na.rm = T)

```

```{r}
sd.now <- filter(sd, as_of_date == "2022-03-01")
sd.mean <- mean(sd.now$percent_of_population_fully_vaccinated, na.rm = TRUE)
sd.mean
```

# Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-02-22”?

```{r}
ggplot(sd.now, aes(percent_of_population_fully_vaccinated))+
  geom_histogram(bins = 35) +
  geom_rug()
```

# Compare UCSD to SD
```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

# Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area:

```{r}
ggplot( ucsd, aes(x= as_of_date, y= percent_of_population_fully_vaccinated)) +
  geom_point() +
  geom_line(group=1) +
  geom_hline(yintercept= sd.mean, col="red", linetype=2)
  labs(x="Date", y="Percent Vaccinated", title = "Vaccination at UCSD")
```


# Subset to all CA areas with a population as large as 92037
```{r}
# Subset to all CA areas with a population as large as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2022-03-01")
```

# Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
sd.36 <- filter(vax.36, as_of_date == "2022-03-01")
sd.36.mean <- mean(sd.now$percent_of_population_fully_vaccinated, na.rm = TRUE)
sd.36.mean
```
```{r}
ggplot(ucsd, aes(x= as_of_date, y= percent_of_population_fully_vaccinated)) +
  geom_point() +
  geom_line(group=1) +
  geom_hline(yintercept= sd.36.mean, col="deep pink", linetype=2) +
  labs(x="Date", y="Percent Vaccinated", title = "Vaccination La Jolla")
```

# Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-02-22”?

```{r}
summary.sd.36 <- summary(sd.36$percent_of_population_fully_vaccinated)
summary.sd.36
```

# Q18. Using ggplot generate a histogram of this data.
```{r}
ggplot(sd.36, aes(percent_of_population_fully_vaccinated))+
  geom_histogram(bins = 35) +
  geom_rug()
```
# Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2022-02-22") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

# Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)
head(vax.36.all)
```

```{r}
ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group=zip_code_tabulation_area) +
  geom_line(alpha=0.2, color= "light blue") +
  ylim(c(0,1)) +
  labs(x="Date", y="Percent Vaccinated",
       title= "Vaccination Rate Across California",
       subtitle="Only areas over 36k") +
  geom_hline(yintercept = 0.75, linetype= 3)
```

# Q21. How do you feel about traveling for Spring Break and meeting for in-person class afterwards?
What Spring Break? Probably best not to travel though. 
