---
title: "Class10"
author: "Morgan Farrell"
date: "2/18/2022"
output:
  pdf_document: default
  html_document: default
---
#Genotype data from 1000 Genomes Project 

Download MXL dataset from https://uswest.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39894595-39895595;v=rs8067378;vdb=variation;vf=105535077#373531_tablePanel

```{r}
mxl <- read.csv("373531-SampleGenotypes.csv")
head(mxl)
```

Find the frequency of G|G homozygous gentoypes in the dataset
```{r}
table(mxl$Genotype..forward.strand.)/ nrow(mxl)
```

Download the GBR dataset
```{r}
gbr <- read.csv("GBR-SampleGenotypes.csv")
head(gbr)
```

Same calculation of G|G with the GBR dataset
```{r}
table(gbr$Genotype..forward.strand.)/ nrow(gbr)
```

# Inital RNA-seq Analysis

## Question 13

Using read.csv first
```{r}
tbl <- read.csv("rs8067378_ENSG00000172057.6.csv", header= TRUE, sep= "")
```

Second method using read.table
```{r}
x <- read.table("rs8067378_ENSG00000172057.6.csv")
head(x)
```

How many of each genoytpe
```{r}
table(x$geno)
```

Find where all the G|G genotypes are
```{r}
x[x$geno == "G/G",]
```

Summary statistics
```{r}
summary(x[x$geno == "G/G",])
```
Check the summary of the other genoytpes
```{r}
#A/G
summary(x[x$geno == "A/G",])
```

```{r}
#A/A
summary(x[x$geno == "A/A",])
```
## Question 14

make a box plot
```{r}
library(ggplot2)
ggplot(x, aes(geno,exp, fill=geno))+
  geom_boxplot(notch=TRUE) +
  geom_jitter(width=0.2, size= 0.5)
```

